Spitzer, Gaia and the Potential of the Milky Way

### 2013-06-25 docs/paper1/paper.tex
paper update

\section{Context for the study}

The method  in Section \ref{measure.sec} takes advantage of three distinct developments in the observations of tidal streams:
(i) the demonstration of a technique for deriving distances to individual RR Lyrae stars with 2\% accuracies (see Section \ref{spitzer.sec});
(ii) the prospect of proper motion measurements of the same stars with $\sim$10 mas/year  precision (see Section  \ref{gaia.sec});
and (iii) the tracing of debris associated with Sgr  entirely around the Galaxy (see Section \ref{sgr.sec})

\subsection{Spitzer RR Lyraes}

\label{spitzer.sec}

%WISE: benedict11? 'precision of spritzer will far exceed these' also 'serendipi…
%Text 'stolen' from proposal 1: the Carnegie RR Lyrae Program

Over twenty-five years ago Longmore, Fernley \& Jameson (1986, MNRAS, 220, 279) demonstrated the existence of a near-infrared period-luminosity (PL) relation for RR
Lyrae stars, using K-band observations of these stars in Galactic globular clusters. Catelan, Pritzl \& Smith (2004, ApJS, 154, 633) showed that a PL relation for RR Lyrae
stars is to be “expected” in the infrared (but vanishingly so in the optical), explained primarily by the changing bolometric correction for these stars as a function of the
observed wavelength. Later, Madore \& Freedman (2011, ApJ, 744, 132) demonstrated the inevitability of decreasing intrinsic scatter with increasing wavelength (and
slope) of PL relations both for Cepheids and for RR Lyrae stars.
Moreover, both reddening effects and metallicity dependencies also decrease significantly for observations further into the IR.

Taken together, these studies suggest observations of RR Lyrae variables by NASA's Spitzer mission could provide us with a stellar distance indicator of unprecedented
accuracy and precision.
\citet{madore13} have demonstrated \citep[using 5 stars with trigonometric parallaxes measured by Hubble][]{benedict11} that the dispersion in the mid-IR PL relation at
Spitzer wavelengths, is on the order of ±0.03 mag. The immediate implication is that it is now possible to determine individual distances that are good to better than ±2\%
for single RR Lyrae stars, out to the limits of Spitzer's ability to detect and measure RR Lyrae variables (which is at least 80 kpc, for modest integration times).

\subsection{Gaia and Spitzer}
\label{gaia.sec}

ESA's Gaia satellite, planned for launch in June 2013, is an astrometric mission which aims to measure the positions of billions of stars with 10-100$\mu$as accuracies
(depending on the color and magnitude of each star), scanning the entire sky multiple times over its 5-year lifetime \citep{perryman01}.
Gaia will revolutionize the field of Galactic Astronomy with sufficient accuracy to create full, 6-dimensional phase-space maps for distances of up to $\sim 10$kpc from the
Sun \citep[see][for discussion]{}.

The left-hand panel of Figure \ref{data.fig} show Gaia's end-of-mission anticipated errors on the distance to  of an individual RR Lyrae star at different distances $d$ from
the Sun.
\footnote{Taken from WWW site?.}
Note that within 2kpc of the Sun, Gaia will be able to measure distances to these stars to better than 2\%: RR Lyraes in this volume will be used to test and calibrate the
Spitzer PL relation.
Beyond 2kpc, the IR PL relation for RR Lyraes will provide more accurate distance estimates than Gaia.

The remaining panels of Figure \ref{data.fig} demonstrate that Gaia's anticipated proper motions errors for RR Lyraes (middle) are sufficiently accurate to distinguish
tangential motions with better than $\sim$10 km/s precision out to many tens of kpc (right-panel).
Hence, the combination of Spitzer and Gaia data offers the exciting prospect of extending the ``horizon'' of where full six-dimensional phase-space maps of our Galaxy are
possible from 10kpc to 80kpc.

%% \begin{figure}
%% \plotone{figs/data.eps}
%% \caption{Summary of data sensitivity: $\Delta d$, $\Delta \mu$ and $\Delta v_{\rm tan}$ as function of $d$.
%% {\it Also put Sgr and Orphan characteristics on this plot?}
%% \label{data.fig}
%% }
%% \end{figure}

\subsection{The Sagittarius debris system}

%% \begin{figure}
%% \plotone{figs/sgr.eps}
%% \caption{
%% \label{sgr.fig}
%% }
%% \end{figure}

@@

\section{Context and motivation}
\subsection{2\% distance errors for RR Lyrae in the halo}

\subsection{Gaia and the age of astrometry}
The Gaia satellite \citep{gaia2001}, planned for launch in October
2013, is an astrometric mission which, amongst a myriad of other
science goals, aims to measure the positions of billions of stars with
10-100~$\mu$as accuracies. With over 50 observations of the entire sky
over the 5-year lifetime of the mission (cite?), Gaia will
revolutionize the field of Galactic Astronomy by providing sufficient
proper motion accuracy to create full, six-dimensional phase-space
maps of the Galaxy for heliocentric distances of up to $\sim$ 10~kpc (xx kpc)
for $V\sim xxx$ ($V\sim yyy$) stars.

@@

\subsection{The Sagittarius debris system}
The Sagittarius Dwarf Galaxy (hereafter Sgr) was discovered
serendipitously during a radial velocity survey of the Galactic bulge
\citep{ibata1994}. While the number of known Milky Way satellites has
more than doubled recently \citep[e.g.,][]{mcconnachie2012}, in many
ways Sgr remains the most studied and most intriguing: it is the
closest satellite, among the largest (only the LMC and SMC are more
luminous), and has the most prominent tidal tails.

The first maps of the Sgr core \citep{ibata1994, ibata1995} revealed
its elongated morphology and suggested tidal deformation, as would be
anticipated from its proximity to the Galaxy. Subsequent N-body models
confirmed this interpretation \citep{velazquez1995} and showed that
debris from Sgr should form coherent streams of stars that encircle
the Galaxy \citep{johnston1995}. Signatures of these [..] stellar
streams have since been discovered in many studies and have been
mapped acorss the sky in carbon stars \citep{totten1998}, M giants
selected from the Two Micron All Sky Survey \citep{majewski2003}, main
sequence turnoff stars from the Sloan Digital Sky Survey
\citep{belokurov2006}, and RR Lyrae in the Catalina Sky Survey
\citep{drake2013}.

@@

Figure \ref{fig:lm10} illustrates results from the \cite[][; hereafter
  LM10]{law2010} N-body simulation of dwarf satellite disruption along
the expected Sgr orbit, with particles projected onto the plane
perpendicular to the Galactic disk containing both the Sun and
Galactic center.  The simulation was run in a potential…..

LM10 demonstrate how a comparison of such simulations with the
observed debris enables a detailed reconstruction of Sgr's history ---
it's mass and stellar content, orbit, rate of destruction and even
original distribution in populations.
(See also other refs ????)
Moreover, the phase-space
distribution of the debris offers a unique probe of the depth, shape
and extent of the Milky Way's dark matter halo that cannot be matched
using other techniques on our own or other galaxies (e.g., Ibata et
al. 2001, ApJ, 551, 294).  Most recently, LM10 have combined all
current data on Sgr debris to assess the triaxiality and orientation
of the outer Galaxy --- the first time that such a reconstruction of
the 3-dimensional mass distribution of a dark matter halo has been
feasible.  These constraints on Sgr's mass and orbit have
invigorated discussions of Sgr in a more cosmological context, the
mass and extent of the original dark matter halo that hosted it as well
as its effect on the MW itself (e.g., Michel-Dansac et
al. 2011, MNRAS, 414, L1).
%\citep{bailin03,purcell11,michel11,gomez12}.

The promise of the combined Spitzer and Gaia measurements of the distances to and proper motions of RR Lyraes in the Sgr debris opens up new approaches to
interpretation.
With 2\% distances measurements, the thickness of Sgr's stream is actually greater
than the uncertainty.  Moreover,
with Gaia's final data release, there will be proper motion
measurements of accuracies less than the intrinsic velocity dispersion for much
of the stream (e.g., corresponding to a tangential velocity accuracy
of 10 km/s for an RR Lyrae at 35kpc) and ground-based radial
velocities of few km/s accuracies can easily be added.
Hence, for the first time, the measured phase-space position of each individual RR Lyrae
relative to the stream is actually physically meaningful.
The next section outlines a method to take full advantage of this accuracy and use each RR Lyrae individually as a potential measure.

\section{Method}
With access to full phase-space information for stars in a tidal
stream, it is no longer necessary to model the orbit of the stream
itself. Instead, it is possible to use the individual stars as
potential measures by exploting the fact that the stars must have come
from the same progenitor. If the present-day orbital parameters of the
progenitor system are known, this places an intuitive constraint on
the potential of the Milky Way: if the orbits of the stars and the
orbit of the progenitor are integrated \emph{backwards} in time, the
stars should `recombine' with the progenitor. If the potential is
incorrect, the orbits of the stars will diverge from that of the
progenitor and thus will not be recaptured by the satellite system.

Quantifying this method requires a sample of stars with known full
space kinematics $(\bs{x}_i, \bs{v}_i)|_{t=0}$, the orbital parameters
for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$, and a
functional form for the potential, $\Phi$\footnote{Future work will
  investigate a more general ... basis function blah blah.}. For a
given set of potential parameters, the orbits of the stars and
progenitor are integrated backwards for some time (determined from an
estimate of the interaction time of the system). The stars and
progenitor system are treated as test particles, so that the debris
stars do not become re-bound to the satellite. Instead, at each
timestep $t_j$, for each particle $i$, a set of normalized, relative
phase-space coordinates, $(\bs{q}_i(t_j), \bs{p}_i(t_j))$, are
computed using the instantaneous tidal radius, escape velocity, and
phase-space position of the progenitor. At a particular timestep,
these coordinates are defined as
\begin{equation}
  \bs{q}_i = \frac{\bs{x}_i -
    \bs{x}_p}{R_{tide}}\,\,\,\,\,\,\,\,\bs{p}_i = \frac{\bs{v}_i -
    \bs{v}_p}{v_{esc}}.
\end{equation}
These definitions require an estimate of the mass of the satellite,
$m_{sat}$, which, combined with the computed enclosed mass of the
potential, $m_{enc}$, sets the instantaneous tidal radius and escape
velocity,
\begin{equation}
  R_{tide}=\Big(\frac{m_{sat}}{3m_{enc}}\Big)^{1/3}\,\,\,\,\,\,\,\,
  v_{esc}=\sqrt{\frac{2Gm_{sat}}{R_{tide}}}.
\end{equation}

Qualitatively, when $D_{ps}=\sqrt{|\bs{q}_i|^2+|\bs{p}_i|^2} <
\sqrt{2}$, the star is probably recaptured by the satellite, however,
this is not imposed as a hard constraint. Instead, we construct a 6D
distribution of the minimum phase-space vectors for each star over the
individual, integrated orbits. For each star, $i$, the phase-space
distance, $D_{ps}$, is computed at each timestep $t_j$, and the vector
with the minimum phase-space distance is stored
\begin{align}
  t^*_i &= \argmin_{t} D_{ps,i}\\
  A_{ik} &= (\bs{q}_i(t^*_i),\bs{p}_i(t^*_i)) % A_ik are really samples from a distribution, or we treat them as such
\end{align}
where $k\in\{1,6\}$. Intuitively, the variance of this distribution
will be larger for orbits integrated in an incorrect potential
relative to the distribution computed from the `true' orbital history
of the stars. Thus, the \emph{generalized variance} of the
distribution, computed for a given set of potential parameters,
$\bs{\theta}$, is a natural choice for the scalar objective function
used in constraining the potential of the Milky Way,
\begin{equation}
  f(\bs{\theta})=\det(\mathrm{Cov}(A_{ik})).
\end{equation}

\subsection{Application to Simulated Data}

>>>>

@@

\section{Context and motivation}
The method presented in Section \ref{sec:method} takes advantage of
three distinct developments in the observations of tidal streams: (i)
the demonstration of a technique for deriving distances to individual
RR Lyrae stars with 2\% accuracies (Section \ref{sec:spitzer}); (ii)
the prospect of proper motion measurements of the same stars with
$\sim$10~mas/yr precision (Section \ref{sec:gaia}); and (iii) the
tracing of debris associated with Sgr around the entire Galaxy
(Section \ref{sec:sgr})

\subsection{2\% distance errors for RR Lyrae in the halo}
\label{sec:spitzer}
Over twenty-five years ago, \citet{longmore1986} demonstrated the
existence of a near-infrared period-luminosity (PL) relation for RR
Lyrae stars using $K$-band observations of these stars in Galactic
globular clusters. More recently, \citet{catelan2004} showed that a PL
relation for RR Lyrae stars is expected in the infrared -- but
vanishingly so in the optical -- and is explained primarily by the
changing bolometric correction for these stars as a function of the
observed wavelength. Later, \citet{madore2012} demonstrated the
inevitability of decreasing intrinsic scatter with increasing
wavelength (and slope) of PL relations both for Cepheids and for RR
Lyrae stars. Both reddening effects and metallicity dependencies also
decrease significantly for observations further into the IR.

These studies suggest that observations of RR Lyrae variables by
NASA's Spitzer mission could provide us with a stellar distance
indicator of unprecedented accuracy and precision. \citet{madore2012}
have demonstrated \citep[using 5 stars with trigonometric parallaxes
  measured by Hubble;][]{benedict2011} that the dispersion in the
mid-IR PL relation--at Spitzer wavelengths--is of order
$\pm$0.03~mag. This implies that it is possible to determine
individual distances that are good to better than $\pm 2\%$ for RR
Lyrae stars out to $\sim$80~kpc (the limit of Spitzer's ability to
detect and measure RR Lyrae variables, for modest integration
times).

\subsection{Gaia and the age of astrometry}
\label{sec:gaia}
The Gaia satellite \citep{gaia2001}, planned for launch in October
2013, is an astrometric mission which, amongst a myriad of other
science goals, aims to measure the positions of billions of stars with
10-100~$\mu$as accuracies. With over 50 observations of the entire sky
over the 5-year lifetime of the mission (cite?), Gaia will
revolutionize the field of Galactic Astronomy by providing sufficient
proper motion accuracy to create full, six-dimensional phase-space
maps of the Galaxy for heliocentric distances of up to $\sim$ 10~kpc (xx kpc)
for $V\sim xxx$ ($V\sim yyy$) stars.

@@

\subsection{The Sagittarius debris system}
\label{sec:sgr}
The Sagittarius Dwarf Galaxy (hereafter Sgr) was discovered
serendipitously during a radial velocity survey of the Galactic bulge
\citep{ibata1994}. While the number of known Milky Way satellites has
more than doubled recently \citep[e.g.,][]{mcconnachie2012}, in many
ways Sgr remains the most studied and most intriguing: it is the
closest satellite, among the largest (only the LMC and SMC are more
luminous), and has the most prominent tidal tails.

The first maps of the Sgr core \citep{ibata1994, ibata1995} revealed
its elongated morphology and suggested tidal deformation, as would be
anticipated from its proximity to the Galaxy. Subsequent N-body models
confirmed this interpretation \citep{velazquez1995} and showed that
debris from Sgr should form coherent streams of stars that encircle
the Galaxy \citep{johnston1995}. Signatures of these [..] stellar
streams have since been discovered in many studies and have been
mapped across the sky in carbon stars \citep{totten1998}, M giants
selected from the Two Micron All Sky Survey \citep{majewski2003}, main
sequence turnoff stars from the Sloan Digital Sky Survey
\citep{belokurov2006}, and RR Lyrae in the Catalina Sky Survey
\citep{drake2013}.

@@

Most recently, \citet[][hereafter LM10]{law2010} have combined all
current data on Sgr debris to assess the triaxiality and orientation
of the outer Galaxy --- the first time that such a reconstruction of
the 3-dimensional mass distribution of a dark matter halo has been
feasible. Figure \ref{fig:lm10} illustrates results from the LM10
N-body simulation of dwarf satellite disruption along the expected Sgr
orbit. The particle positions are shown in a coordinate system defined
by the orbit of the satellite orbit \citep{majewski2004}. The
simulation was run in a three-component potential, with a triaxial,
logarithmic halo model of the form
\begin{equation}
  \Phi_{halo} = v^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + R_c^2).
\end{equation}

LM10 demonstrate how a comparison of such simulations with the
observed debris enables a detailed reconstruction of Sgr's history ---
it's mass and stellar populations, orbit, rate of destruction, and
even original distribution in populations (other refs??). Moreover,
the phase-space distribution of the debris offers a unique probe of
the depth, shape, extent, and orientation of the Milky Way's dark
matter halo that cannot be matched using other techniques on our own
or other galaxies \citep[e.g.,][]{ibata2001}. The LM10 constraints on
Sgr's mass and orbit have invigorated discussions of Sgr in a more
cosmological context, the mass and extent of the original dark matter
halo that hosted it, as well as its effect on the Milky Way itself
\citep{bailin2003,purcell2011,micheldansac2011,gomez2012}.

The combined Spitzer and Gaia measurements of distances and proper
motions of RR Lyraes in the Sgr debris opens up new avenues of
interpretation. A 2\% distance error at $\sim$30~kpc is smaller than
the predicted thickness of the stream, ranging from
$\sim2-10$~kpc. Combined with Gaia tangential velocity measurement
accuracies ($\sim$2~km/s at 30~kpc) less than the intrinsic dispersion
for much of the stream ($\sigma_v\sim ??$), and ground-based radial
velocities with $\sim10$~km/s accuracy, the measured phase-space
position of each individual RR Lyrae relative to the stream becomes a
powerful probe of the potential: the next section outlines a new
method to take advantage of this information and use each star as an
individual potential measure.

\section{Method}
\label{sec:method}
With access to full phase-space information for stars in a tidal
stream, it is no longer necessary to model the orbit of the stream
itself. Instead, it is possible to use the individual stars as
potential measures by exploting the fact that the stars must have come
from the same progenitor. If the present-day orbital parameters of the
progenitor system are known, this places an intuitive constraint on
the potential of the Milky Way: if the orbits of the stars and the
orbit of the progenitor are integrated \emph{backwards} in time, the
stars should `recombine' with the progenitor. If the potential is
incorrect, the orbits of the stars will diverge from that of the
progenitor and thus will not be recaptured by the satellite system.

Quantifying this method requires a sample of stars with known full
space kinematics $(\bs{x}_i, \bs{v}_i)|_{t=0}$, the orbital parameters
for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$, and a
functional form for the potential, $\Phi$.\footnote{Future work will
  investigate a more general ... basis function blah blah.} For a
given set of potential parameters, the orbits of the stars and
progenitor are integrated backwards for some time (determined from an
estimate of the interaction time of the system). The stars and
progenitor system are treated as test particles, so that the debris
stars do not become re-bound to the satellite. Instead, at each
timestep $t_j$, for each particle $i$, a set of normalized, relative
phase-space coordinates, $(\bs{q}_i(t_j), \bs{p}_i(t_j))$, are
computed using the instantaneous tidal radius, escape velocity, and
phase-space position of the progenitor. At a particular timestep,
these coordinates are defined as
\begin{equation}
  \bs{q}_i = \frac{\bs{x}_i -
    \bs{x}_p}{R_{tide}}\,\,\,\,\,\,\,\,\bs{p}_i = \frac{\bs{v}_i -
    \bs{v}_p}{v_{esc}}
\end{equation}
where $\bs{x}_i$ and $\bs{v}_i$ are the 3D position and velocity of
the $i$th particle, and $\bs{x}_p$ and $\bs{v}_p$ are the same for the
satellite.  These definitions require an estimate of the mass of the
satellite, $m_{sat}$, which, combined with the computed enclosed mass
of the potential, $m_{enc}$, sets the instantaneous tidal radius and
escape velocity,
\begin{equation}
  R_{tide}=\Big(\frac{m_{sat}}{3m_{enc}}\Big)^{1/3}\,\,\,\,\,\,\,\,
  v_{esc}=\sqrt{\frac{2Gm_{sat}}{R_{tide}}}.
\end{equation}

Qualitatively, when the distance in this normalized 6D space,
$D_{ps}=\sqrt{|\bs{q}_i|^2+|\bs{p}_i|^2} < \sqrt{2}$, the star is
probably recaptured by the satellite (but this is not imposed as a
hard constraint). As a statistical measure over all stars, we
construct a 6D distribution of the minimum phase-space vectors for
each star over the individual, integrated orbits. For each star, $i$,
the phase-space distance, $D_{ps}$, is computed at each timestep
$t_j$, and the vector with the minimum phase-space distance is stored
\begin{align}
  t^*_i &= \argmin_{t} D_{ps,i}\\
  A_{ik} &= (\bs{q}_i(t^*_i),\bs{p}_i(t^*_i)) % A_ik are really samples from a distribution, or we treat them as such
\end{align}
where $k\in\{1,6\}$. Intuitively, the variance of this distribution
will be larger for orbits integrated in an incorrect potential
relative to the distribution computed from the `true' orbital history
of the stars; in an incorrect potential, small differences in the
orbits of the stars relative to the orbit of the progenitor cause them
to spread out in this relative, normalized phase space. Thus, the
\emph{generalized variance} of the distribution, computed for a given
set of potential parameters, $\bs{\theta}$, is a natural choice for
the scalar objective function used in constraining the potential of
the Milky Way,
\begin{equation}
  f(\bs{\theta})=\det(\mathrm{Cov}(A_{ik})).
\end{equation}

\subsection{Application to Simulated Data}
The LM10 (see Section \ref{sec:sgr}) simulation data is a perfect
test-bed for evaluating the effectiveness of this method. We start by
extracting both particle data and the satellite orbital parameters
from the present-day snapshot of the simulation
data.\footnote{www.astro.virginia.edu/~srm4n/Sgr/data.html} We then
`observe' a sample of 100 stars from the first leading and trailing
wraps of the stream: the radial velocity errors are drawn from a
Gaussian $\epsilon_{rv} \sim \mathcal{N}(\mu=0,\sigma=10~{\rm km/s})$,
the tangential velocity errors depend on the distance to the star and
are computed from the expected Gaia error curve (shown in
Figure~\ref{fig:gaia}), and the distance errors on a star of distance
$D$ are $\epsilon_{dist} \sim \mathcal{N}(0,0.02\times
D)$. Figure~\ref{fig:} shows all 2D projections of the 6D minimum
phase-space distance distribution for the observed particles in the
correct and an incorrect potential. Visually, it is clear that the
variance of the distribution increases in the wrong potential. The
generalized variance defines a convex hypersurface over which we
optimize four of the six logarithmic potential parameters:
$v_{circ}$, $\phi$, $q_1$, and $q_z$. Figure~\ref{fig:} shows [...]
curves produced by varying the true parameters of the potential by
10\% around the true value.
